---
title: "Analysis of Setty et al."
subtitle: "Data Preparation"
author: "Priyansh Srivastava"
date: "`r Sys.Date()`"
output:
    ioslides_presentation:
        widescreen: false
        incremental: true
        css: css/Data_Preparation.css
                
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
suppressPackageStartupMessages(library(shiny))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(rhdf5))
```

## <span style="color: #000000;">Outline</span>

- <b>Brief description of the study.</b>

- <b>Methods,</b>
<ul>
  <li>About the dataset</li>
  <li>Sequencing</li>
</ul>

- <b>Data Exploration,</b>
<ul>
  <li>How to download data</li>
  <li>Basic Exploration</li>
</ul>

- <b>Trajectory Analysis,</b>
<ul>
  <li>From the Article</li>
  <li>Monocle3</li>
  <li>Slingshot</li>
</ul>

## <span style="color: #000000;">About the Study</span>

<p class="p_tags"><strong>Title:</strong> Characterization of cell fate probabilities in single-cell data with Palantir.</p>

<p class="p_tags"><strong>Authors:</strong> Manu Setty, Vaidotas Kiseliovas, Jacob Levine, Adam Gayoso, Linas Mazutis & Dana Pe'er.</p>

<p class="p_tags"><strong>Journal:</strong> Nature Biotechnology.</p>

<p class="p_tags"><strong>Date/Year:</strong> $18^{th}$ September 2019</p>

<p class="p_tags"><strong>Main Ideas:</strong></p>

- <p class="bullets">Addressed discrete versus continuous nature of both differentiation and cell fate.</p>
- <p class="bullets">Present an algorithm "<strong>Palantir</strong>" that models trajectories of differentiating cells by treating cell fate as a probabilistic process and leverages entropy to measure cell plasticity along the trajectory.</p>
- <p class="bullets">Performance of Palantir "<strong>human bone marrow single-cell RNA sequencing data</strong>".</p>
- <p class="bullets">Comparison of Palantir with pre-existing trajectory inference tools.</p>

# About the dataset

## <span style="color: #000000;">Experimental Setup</span>


```{r, echo=FALSE, "Data-prep"}
# Read raw metadata
metadata1 <- read.csv("../datasets/setty_et_al/metadata.tsv", sep = "\t")
metadata2 <- read.csv("../datasets/setty_et_al/cell_type.csv", sep = ",")
metadata <- merge(metadata2, metadata1, "cell_suspension.biomaterial_core.biomaterial_id")

# Select the columns of Interest
metadata.clean <- metadata[,c("cell_suspension.biomaterial_core.biomaterial_id",
                              "annotated_cell_identity.text",
                              "annotated_cell_identity.ontology",
                              "barcode", "donor_organism.sex",
                              "donor_organism.biomaterial_core.biomaterial_id",
                              "donor_organism.organism_age") ,drop = F]

# Independently dividing replicates
metadata.rep1 <- metadata.clean[metadata.clean$donor_organism.biomaterial_core.biomaterial_id == "HS_BM_1", ]
metadata.rep2 <- metadata.clean[metadata.clean$donor_organism.biomaterial_core.biomaterial_id == "HS_BM_2", ]
metadata.rep3 <- metadata.clean[metadata.clean$donor_organism.biomaterial_core.biomaterial_id == "HS_BM_3", ]

# Remove duplicates if any
metadata.rep1 <- metadata.rep1[!duplicated(metadata.rep1),]
metadata.rep2 <- metadata.rep2[!duplicated(metadata.rep2),]
metadata.rep3 <- metadata.rep3[!duplicated(metadata.rep3),]

# rename list
metadata.list <- list(metadata.rep1 = metadata.rep1, 
                      metadata.rep2 = metadata.rep2,
                      metadata.rep3 = metadata.rep3)

# Renaming columns 
metadata.list <- lapply(metadata.list, function(X){
    colnames(X) <- c("biomaterial_id", "cell_identity", "cell_identity_ontology",
                     "barcode", "sex", "id", "age")
        return(X)
})
```

## <span style="color: #000000;">CD34$^{+}$ Human Bone Marrow Cells</span>

- <p class="bullets2">Cryopreserved bone marrow stem/progenitor CD34$^{+}$ cells from healthy donors were purchased from <strong>AllCells LLC</strong>.</p>
\
- <p class="bullets2">scRNA-Seq was performed with <strong>10X</strong> Genomics System.</p>
\
- <p class="bullets2"><strong>Single Cell 3'</strong> Library and <strong>Gel Bead kit V2</strong> was used for library preparation.</p>
\
- <p class="bullets2">Briefly, <strong>8,700 cells</strong> (viability 90–97%) were loaded per reaction, targeting recovery of <strong>5,000 cells</strong> with <strong>3.9% multiplet rate</strong>.</p>
\
- <p class="bullets2"><strong>12 cycles of PCR amplification, per cell</strong>.</p>
\
- <p class="bullets2">Sequenced on an <strong>Illumina HiSeq 2500 system</strong> (HiSeq SBS V4 chemistry kit).</p>

---


```{r, echo=FALSE}
# Clean the Metadata file
shinyApp(
  ui = fluidPage(
    radioButtons(inputId = "replicate", inline = TRUE,
                 label = "Select Replicate:",
                 choices = c("Donor-1" = "metadata.rep1",
                             "Donor-2" = "metadata.rep2",
                             "Donor-3" = "metadata.rep3")
                 ),
    p("-Number of Cells", span(textOutput(outputId = "nCells", inline=T), style="color:blue")),
    p("-Age of Donor", span(textOutput(outputId = "age", inline=T), style="color:blue")),
    p("-Sex of Donor", span(textOutput(outputId = "sex", inline=T), style="color:blue")),
    h5("Cell types"), plotOutput("cellTypesPlot", width = "100%",height = "300px")
  ),
  server = function(input, output) {
      # Number of Cells
      output$nCells = renderText({
          nCells <- nrow(metadata.list[[input$replicate]])
          return(nCells)
          })
      
      # Age of Donor
      output$age = renderText({
          age <- unique(metadata.list[[input$replicate]][["age"]])
          return(age)
          })
      
      # Age of Donor
      output$sex = renderText({
          sex <- unique(metadata.list[[input$replicate]][["sex"]])
          return(sex)
          })
      
      # Cell types
      output$cellTypesPlot = renderPlot({
          cellTypesPlotData <- metadata.list[[input$replicate]]
          cellTypesPlot <- ggplot(cellTypesPlotData,
                                  aes(x=cell_identity,  fill = cell_identity)) + geom_bar() + theme_classic()  + theme(legend.position = "none", axis.text.x = element_text(size = 12, angle = 45,vjust = 1, hjust=1), axis.text.y = element_text(size = 12), axis.title = element_text(size = 12)) + xlab("Cell Types") +ylab("Number of Cells") + coord_flip()
          
          return(cellTypesPlot)
          })
      
  })

```

```{r}

h5ls("datasets/setty_et_al/human_cd34_bm_rep1.h5ad")

data = h5read("datasets/setty_et_al/human_cd34_bm_rep1.h5ad","/raw.X/data")
indices = h5read("datasets/setty_et_al/human_cd34_bm_rep1.h5ad","/raw.X/indices")
indptr = h5read("datasets/setty_et_al/human_cd34_bm_rep1.h5ad","/raw.X/indptr")

obs <-  h5read("datasets/setty_et_al/human_cd34_bm_rep1.h5ad","/obs")
obs <-  h5read("datasets/setty_et_al/human_cd34_bm_rep1.h5ad","/raw.var")

```




